$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-01 paper/appendix.tex
Polish and shorten the section on multiplicities

%% Appendix
\section{Structural Congruence}
\label{app:struct}
\begin{nilemma}
  \label{lm:lower-lift}
  The function $lower_i \; P \; uP$ has an inverse $lift_i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $lift_i \; (lower_i \; P \; uP) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

>>>>

%% Appendix
\section{Structural Congruence}
\label{app:struct}

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen anywhere inside a process, and are closed under reflexivity, symmetry and transitivity as shown by the first row of \autoref{fig:struct-cong1}.
The rest of the rules defines structural congruence under a context $\mathcal{C}[\cdot]$ \cite{Sangio01}, respectively restriction, composition, input and output.

@@

\section{Usage Algebra}
\label{app:usage-algebra}

\todo{we haven't introduced the forall and implicit syntax}
\begin{nidefinition}
  The algebra on multiplicities must be \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}, and needs a \emph{minimal neutral element}.
  It also requires a special element $\one$ that is to be used to count inputs and outputs.
  We define the laws of such a \emph{partial commutative monoid} as a type $\Quantifier_C$ on a carrier set $C$ in \autoref{fig:multiplicities} --- we use a terniary relation to model the monoidal operation.

  \begin{figure}[h]
  \begin{equation}
  \begin{aligned}
    &\zero                  &:{} &                      &        & C \\
    &\one                   &:{} &                      &        & C \\
    &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to et \\
    &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
    &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\field{0\cdot-min^l}   &:{} &\forall \{y z\}       & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\field{\cdot-id^l}     &:{} &\forall \{x\}         & \to \; & \op{x}{\zero}{x} \\
    &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation}
  \caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid.}
  \label{fig:multiplicities}
  \end{figure}
\end{nidefinition}

\section{Lemmas for Type Safety}
\label{app:lemmas-type-safety}

\begin{nilemma}
  \label{lm:lower-lift}
  The function $lower_i \; P \; uP$ has an inverse $lift_i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $lift_i \; (lower_i \; P \; uP) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

### 2020-05-01 paper/paper.tex
Polish and shorten the section on multiplicities

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
We increase the \emph{depth} counter by 1 in channel restriction and input processes.
\autoref{fig:syntax} showcases a \picalc{} syntax that is well-scoped by construction thanks to type-level de Bruijn indices.

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} indexed by the channel $i$ on which communication occurs (\autoref{fig:reduction}).
In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel ${i}$.
As a result of that communication, the continuation of the input process $P$ gets all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After substitution, $\subst{P}{j}{0}$ is \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input or output, otherwise the sequencing of actions would not be preserved \cite{Sangio01}.

@@

A resource-aware typing system for the \picalc{} must take resource usages into account, and needs an algebra on these --- a way of \emph{splitting} resources on parallel composition, of \emph{ticking off} resources on input and output.
We characterise such an algebra in  \autoref{multiplicities}.
We define independent typing contexts and usage contexts in  \autoref{contexts}.
We provide a typing system with leftovers for a resource-aware \picalc{} in  \autoref{leftover-typing}.

@@

In the linear \picalc{} each channel has an input or an output \emph{capability}, and each capability has a given \emph{multiplicity} of 0 (exhausted) or 1 (available).
We generalise multiplicities by defining an algebra such that linear, gradual and shared types all satisfy these algebraic laws.

\todo{we haven't introduced the forall and implicit syntax}
\begin{nidefinition}
  The algebra on multiplicities must be \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}, and needs a \emph{minimal neutral element}.
  It also requires a special element that is to be used to count inputs and outputs.
  We define the laws of such a \emph{partial commutative monoid} as a type $\Quantifier_C$ on a carrier set $C$ in \autoref{fig:multiplicities} --- we use a terniary relation to model the monoidal operation.

  \begin{figure}[h]
  \begin{equation}
  \begin{aligned}
    &\zero                  &:{} &                      &        & C \\
    &\one                   &:{} &                      &        & C \\
    &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to et \\
    &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
    &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\field{0\cdot-min^l}   &:{} &\forall \{y z\}       & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\field{\cdot-id^l}     &:{} &\forall \{x\}         & \to \; & \op{x}{\zero}{x} \\
    &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation}
  \caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid.}
  \label{fig:multiplicities}
  \end{figure}
\end{nidefinition}

\begin{note}
  We will often work with input and output multiplicities as pairs.
  We use the notation $\Carrier^2$ to stand for a $\Carrier \times \Carrier$ tuple, and $\op{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where all properties are trivially lifted.
  Henceworth, we use $\lz$ to denote the pair $(0 \comma 0)$, $\li$ for the pair $(1 \comma 0)$, $\lo$ for $(0 \comma 1)$ and $\lio$ for $(1 \comma 1)$, borrowing from the linear \picalc{} notation \cite{KPT96,Sangio01}.
\end{note}

Linear, gradual and shared types are all defined as an instance of the laws in \autoref{fig:multiplicities}.
We illustrate these types in typing derivations in \autoref{example-derivations}.

\begin{description}
  \item [Linear]
    The carrier $C$ is implemented as a type $\type{Two}$ with the trivial elements $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{\one}$), and is uninhabited in every other case.
    All other rules follow trivially.

    \item [Gradual]
    The carrier $C$ is implemented as the type of natural numbers $\N$.
    The element $\zero$ corresponds to 0, and $\one$ to 1.
    The partial monoid $\op{x}{y}{z}$ is defined exactly when $x \equiv y + z$.
    All other rules follow trivially from the algebraic rules for the addition of natural numbers.

    \item [Shared]
    The carrier $C$ is a type $\type{One}$ with a single trivial constructor $\constr{\omega}$.
    Both $\zero$ and $\one$ are instantiated to $\constr{\omega}$.
    The relation $\op{\constr{\omega}}{\constr{\omega}}{\constr{\omega}}$ is inhabited.
    All other rules follow trivially.
\end{description}

@@

  \begin{figure}[h]
  \begin{equation}
  \begin{split}
    &\field{IDX}          &:{} &et \\
    &\field{\exists IDX}  &:{} &\field{IDX} \\
    &\field{CARRIER}      &:{} &\field{IDX} \to et \\
    &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
  \end{split}
  \end{equation}
  \caption{Indexed set of partial commutative monoids.}
  \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

>>>>

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
We increase the \emph{depth} counter by 1 in channel restriction and input processes.
\autoref{fig:syntax} showcases a \picalc{} syntax that is well-scoped by construction using type-level de Bruijn indices.

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} indexed by the channel $i$ on which communication occurs (\autoref{fig:reduction}).
In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel ${i}$.
As a result of that communication, the continuation of the input process $P$ gets all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After substitution, $\subst{P}{j}{0}$ is \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

@@

We characterise a usage algebra for our typing system in \autoref{multiplicities}.
This algebra defines the way in which resources are \emph{split} on parallel composition; the way in which resources are \emph{consumed} on input and output.
We then define typing and usage contexts in \autoref{contexts}.
We provide a typing system for a resource-aware \picalc{} in \autoref{leftover-typing}.

@@

In the linear \picalc{} each channel has an input and an output \emph{capability}, and each capability has a given \emph{multiplicity} of 0 (exhausted) or 1 (available).
We generalise over this notion by defining an algebra that is satisfied by linear, gradual and shared types.

\begin{nidefinition}
  Our usage algebra is a ternary relation $\op{x}{y}{z}$ that is \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
  It has an element $\one$ that is used to count inputs and outputs.
  For a formal definition, refer to \autoref{fig:multiplicities} in \autoref{app:usage-algebra}.
\end{nidefinition}

\begin{note}
  We will often work with pairs of input and output multiplicities.
  We use the notation $\Carrier^2$ to stand for a $\Carrier \times \Carrier$ tuple of input and output multiplicities, respectively.
  We use $\op{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where the algebraic laws are trivially lifted.
  Henceworth, we use $\lz$ to denote the pair $(0 \comma 0)$, $\li$ for the pair $(1 \comma 0)$, $\lo$ for $(0 \comma 1)$ and $\lio$ for $(1 \comma 1)$, borrowing from the linear \picalc{} notation \cite{KPT96,Sangio01}.
\end{note}

Linear, gradual and shared types are all defined as an instance of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivations}.

\begin{description}
  \item [Linear]
    The carrier is a type $\type{Two}$ with the trivial elements $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{\one}$), and is uninhabited in every other case.
    All other rules follow trivially.

    \item [Gradual]
    The carrier is the type of natural numbers $\N$.
    The element $\zero$ corresponds to 0, and $\one$ to 1.
    The partial monoid $\op{x}{y}{z}$ is defined exactly when $x \equiv y + z$.
    All other rules follow trivially from the algebraic rules for the addition of natural numbers.

    \item [Shared]
    The carrier is a type $\type{One}$ with a single trivial constructor $\constr{\omega}$.
    Both $\zero$ and $\one$ are instantiated to $\constr{\omega}$.
    The relation $\op{\constr{\omega}}{\constr{\omega}}{\constr{\omega}}$ is inhabited.
    All other rules follow trivially.
\end{description}

@@

  \begin{figure}[h]
  \begin{equation}
  \begin{split}
    &\field{IDX}          &:{} &\Set \\
    &\field{\exists IDX}  &:{} &\field{IDX} \\
    &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
    &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
  \end{split}
  \end{equation}
  \caption{Indexed set of partial commutative monoids.}
  \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}
